Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-26 article.tex
wordings

We report here experiments on the \textit{normal}-1,2,3, and 4
datasets provided by the organizers of the Connectomics
challenge. Each of these datasets is obtained from the simulation
\citep{stetter2012model} of different neural networks of 1000 neurons
and about 15000 edges each (i.e., a network density of about
1.5\%). Each neuron is described by a calcium fluorescence time series
of length $T=179500$. All inference methods compared here provide a
ranking of all pairs of neurons according to some association
score. To assess the quality of this ranking, we compute both ROC and
precision-recall curves against the ground truth network, which are
summarized using the area under the curve, denoted respectively AUROC
and AUPRC. Only the AUROC score was used to rank the challenge
participants but the precision-recall curve has been shown to be a
more sensible metric for network inference, especially when network
density is small (see e.g. \cite{schrynemackers2013protocols}).
Since neurons are not self-connected in the ground truth networks
(i.e., $(i, i) \not \in E, \forall i \in V$), we have manually set the
score of such edges to the minimal possible association score before
computing ROC and PR curves.

@@

The top of Table \ref{tab:comparison} reports AUROC and AUPRC for all
four networks using in each case partial correlation with different
filtering functions. Except for the last two rows that uses PCA, the
exact inverse of the covariance matrix was used in each case. These
results clearly show the importance of the filters. AUROC increases in
average from 0.77 to 0.93. PCA does not really affect AUROC scores but
it improves significantly AUPRC scores. Taking the average over
various parameter setting gives an improvement of 10\% in AUPRC but
only a minor change in AUROC. The last row shows the final performance
of the method specifically tuned for the challenge (see the Appendix
\ref{app:optimized} for all details). Although this tuning was
decisive to obtain the best performance at the challenge, it does not
significantly improve neither AUROC nor AUPRC.

\begin{table}[t]\label{tab:comparison}
%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Method} $\backslash$ \textit{normal-} & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
\hline
\hline
No  filtering       					& 0.777 & 0.767 & 0.772 & 0.774 & 0.070 & 0.064 & 0.068 & 0.072\\
$ h \circ g \circ f_1$                  & 0.923 & 0.925 & 0.923 & 0.922 & 0.311 & 0.315 & 0.313 & 0.304\\
$ w \circ h \circ g \circ f_1$          & 0.931 & 0.929 & 0.928 & 0.926 & 0.326 & 0.323 & 0.319 & 0.303\\
+ PCA         							& 0.932 & 0.930 & 0.928 & 0.926 & 0.355 & 0.353 & 0.350 & 0.333\\
Averaging           					& 0.938 & 0.936 & 0.936 & 0.932 & 0.391 & 0.389 & 0.385 & 0.374\\
Full method           					& 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388\\
\hline
Correlation & 0.886 & 0.884 & 0.891 &  0.877 & 0.153 & 0.145 & 0.170 & 0.132\\
GTE & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142\\
GENIE3 & & & & & & & & \\
\hline
\end{tabular}
\caption{Top: ROC (left) and PR (right) AUC on \text{normal-1,2,3,4} with partial correlation and different filtering functions.
Bottom: ROC (left) and PR (right) AUC on \text{normal-1,2,3,4} with different methods.}
\end{table}

@@

In the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalized transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:curves}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer of the
DREAM5 challenge \citep{marbach2012}. When transposed to neural networks, this
method uses as a confidence score for the edge going from neuron $i$ to neuron
$j$ the importance score of variable $X_i$ in a Random Forests model trying to
predict $X_j$ from all variables in $X\setminus X_j$. To reduce the
computational cost of this method, we had to limit each tree in the Random
Forests model to a maximum depth of 3. This constraint potentially severly
affects the performance of this method with respect to the use of fully grown
trees. We nevertheless provide these results for comparison purpose. PC and
GENIE3 were both applied on the same pre-filtered time series ($g \circ
f_1$). For GENIE3, we built 10000 trees per neuron and we used default settings
for all other parameters (except for the maximal tree depth). For GTE, we
reproduced the exact same setting (conditioning level and pre-processing) that
was used for the challenge.

@@

\begin{figure}[t]
\centering
\subfigure[ROC curves]{\includegraphics[width=0.45\textwidth]{images/curve_roc} \label{fig:roc_curve}}
\subfigure[Precision-recall curves]{\includegraphics[width=0.45\textwidth]{images/curve_pr} \label{fig:pr_curve}}
\caption{ROC (left) and PR (right) curves on \text{normal-2} with different methods. Areas under the curves are reported in the legend.}
%% The weighted average of partial correlation matrix, winning approach
         %% of the challenge, yields better than AUROC and AUPRC than all
         %% other state-of-the-art methods. (parameters: $\tau=0.11$ except for
         %% averaged partial correlation method, partial correlation used
         %% filter $f_1$). Area under the curves are given between $()$
         %% for the corresponding methods.
%}
\label{fig:curves}
\end{figure}

>>>>

We report here experiments on the \textit{normal-1,2,3}, and \textit{4}
datasets provided by the organizers of the Connectomics
challenge. Each of these datasets is obtained from the simulation
\citep{stetter2012model} of different neural networks of 1000 neurons
and about 15000 edges each (i.e., a network density of about
1.5\%). Each neuron is described by a calcium fluorescence time series
of length $T=179500$. All inference methods compared here provide a
ranking of all pairs of neurons according to some association
score. To assess the quality of this ranking, we compute both ROC and
precision-recall curves against the ground truth network, which are
summarized using the area under the curve and respectively  denoted AUROC
and AUPRC. Only the AUROC score was used to rank the challenge
participants but the precision-recall curve has been shown to be a
more sensible metric for network inference, especially when network
density is small (see e.g., \cite{schrynemackers2013protocols}).
Since neurons are not self-connected in the ground truth networks
(i.e., $(i, i) \not \in E, \forall i \in V$), we have manually set the
score of such edges to the minimal possible association score before
computing ROC and PR curves.

@@

The top of Table \ref{tab:comparison} reports AUROC and AUPRC for all
four networks using in each case partial correlation with different
filtering functions. Except for the last two rows that use PCA, the
exact inverse of the covariance matrix was used in each case. These
results clearly show the importance of the filters. AUROC increases in
average from 0.77 to 0.93. PCA does not really affect AUROC scores but
it improves significantly AUPRC scores. Taking the average over
various parameter setting gives an improvement of 10\% in AUPRC but
only a minor change in AUROC. The last row shows the final performance
of the method specifically tuned for the challenge (see the Appendix
\ref{app:optimized} for all details). Although this tuning was
decisive to obtain the best performance at the challenge, it does not
significantly improve neither AUROC nor AUPRC.

\begin{table}[t]\label{tab:comparison}
%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Method} $\backslash$ \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
No  filtering       					& 0.777 & 0.767 & 0.772 & 0.774 & 0.070 & 0.064 & 0.068 & 0.072\\
$ h \circ g \circ f_1$                  & 0.923 & 0.925 & 0.923 & 0.922 & 0.311 & 0.315 & 0.313 & 0.304\\
$ w \circ h \circ g \circ f_1$          & 0.931 & 0.929 & 0.928 & 0.926 & 0.326 & 0.323 & 0.319 & 0.303\\
+ PCA         							& 0.932 & 0.930 & 0.928 & 0.926 & 0.355 & 0.353 & 0.350 & 0.333\\
Averaging           					& 0.938 & 0.936 & 0.936 & 0.932 & 0.391 & 0.389 & 0.385 & 0.374\\
Full method           					& 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388\\
\hline
PC & 0.886 & 0.884 & 0.891 &  0.877 & 0.153 & 0.145 & 0.170 & 0.132\\
GTE & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142\\
GENIE3 & & & & & & & & \\
\hline
\end{tabular}
\caption{Top: Performance on \textit{normal-1,2,3,4} with partial correlation and different filtering functions.
Bottom: Performance on \textit{normal-1,2,3,4} with different methods.}
\end{table}

@@

In the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalized transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:curves}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer of the
DREAM5 challenge \citep{marbach2012}. When transposed to neural networks, this
method uses as a confidence score for the edge going from neuron $i$ to neuron
$j$ the importance score of variable $X_i$ in a Random Forest model trying to
predict $X_j$ from all variables in $X\setminus X_j$. To reduce the
computational cost of this method, we had to limit each tree in the Random
Forest model to a maximum depth of 3. This constraint potentially severly
affects the performance of this method with respect to the use of fully grown
trees. We nevertheless provide these results for comparison purpose. PC and
GENIE3 were both applied on the same pre-filtered time series ($g \circ
f_1$). For GENIE3, we built 10000 trees per neuron and we used default settings
for all other parameters (except for the maximal tree depth). For GTE, we
reproduced the exact same setting (conditioning level and pre-processing) that
was used for the challenge.

@@

\begin{figure}[t]
\centering
\subfigure[ROC curves]{\includegraphics[width=0.45\textwidth]{images/curve_roc} \label{fig:roc_curve}}
\subfigure[Precision-recall curves]{\includegraphics[width=0.45\textwidth]{images/curve_pr} \label{fig:pr_curve}}
\caption{ROC (left) and PR (right) curves on \textit{normal-2} for the compared methods. Areas under the curves are reported in the legend.}
%% The weighted average of partial correlation matrix, winning approach
         %% of the challenge, yields better than AUROC and AUPRC than all
         %% other state-of-the-art methods. (parameters: $\tau=0.11$ except for
         %% averaged partial correlation method, partial correlation used
         %% filter $f_1$). Area under the curves are given between $()$
         %% for the corresponding methods.
%}
\label{fig:curves}
\end{figure}
